% Section 1: Introduction

\section{Introduction}

Tabular data remains one of the most common data modalities in applied machine learning, appearing in domains such as materials science, healthcare, e-commerce, and human-resource analytics. Despite the rapid progress of deep learning on images, language, and sequential data, tabular learning is still difficult because useful signals are often sparse, heterogeneous, and distributed across mixed numerical and categorical variables. Prior benchmarking studies show that gradient-boosted decision trees (GBDTs), including XGBoost and LightGBM, remain difficult to outperform on many typical tabular datasets because their axis-aligned splits, robustness to feature scale, and ability to model non-linear interactions match the structure of many tabular problems~\citep{chen2016xgboost,ke2017lightgbm,grinsztajn2022tree,holzmuller2024better}.

This project replicates and evaluates xRFM, a recent feature learning approach for tabular data based on the average gradient outer product (AGOP)~\citep{beaglehole2025xrfm,radhakrishnan2024mechanism}. Instead of relying only on fixed input features, xRFM uses gradient-based information from a learned prediction function to estimate directions in the input space that are most relevant to the target. This produces both a predictive model and an interpretable feature-relevance signal. The central question of our replication is therefore whether AGOP-driven feature learning can offer a competitive alternative to strong tabular baselines, especially when compared with XGBoost, LightGBM, and a multilayer perceptron (MLP).

We evaluate all models under a unified experimental pipeline across five public tabular datasets: Diamonds, Superconductivity, Online Shoppers Purchasing Intention, Stroke Prediction, and HR Attrition. These datasets cover regression and binary classification, large- and small-sample regimes, low- and high-dimensional feature spaces, and mixed feature types. The report is organised as follows. Section~2 describes the xRFM algorithm, the AGOP interpretation, dataset preprocessing, evaluation metrics, and model configurations. Section~3 compares predictive performance, training efficiency, scalability, and interpretability. Section~4 discusses when xRFM appears useful and when it is clearly dominated by tree-based methods. Section~5 summarises the main findings and future directions.

Overall, our results show that tree-based models remain the strongest general-purpose baselines for these tabular tasks. xRFM is competitive on selected small-scale classification settings, particularly HR Attrition, but underperforms substantially on large-scale regression such as Diamonds. This suggests that AGOP-based feature learning is promising as an interpretable modelling strategy, but it requires careful tuning and may not yet replace highly optimised GBDT pipelines for general tabular prediction.
