% Section 5: Conclusion

\section{Conclusion}

In this project, we replicated xRFM, an AGOP-driven feature learning method for tabular data, and conducted a reproducible comparative evaluation against XGBoost, LightGBM, and MLP across five public datasets covering regression and binary classification. A unified pipeline was used for preprocessing, splitting, training, and evaluation, reducing the risk of cross-model discrepancies in data usage and metric calculation.

The experimental findings are threefold. First, tree-based gradient boosting models remain the strongest general-purpose baselines for the evaluated tabular tasks, with LightGBM and XGBoost providing the best overall balance of predictive accuracy and efficiency. Second, xRFM shows selective promise: it is competitive on Superconductivity and achieves the highest accuracy on HR Attrition, suggesting that AGOP-based feature learning can be useful in some small-data or numerical feature-learning settings. However, it underperforms substantially on Diamonds and scales poorly on that dataset, indicating that AGOP feature learning is not automatically superior to tree-based splitting on large mixed-type tabular regression. Third, MLP performance is less stable, especially under class imbalance and large-scale tabular regression.

The interpretability analysis shows that xRFM provides meaningful AGOP-diagonal feature rankings, such as size-related variables for Diamonds, \texttt{PageValues} for Online Shoppers, and overtime-related variables for HR Attrition. These rankings are useful as internal explanations of xRFM's learned feature geometry, but they should be interpreted alongside external methods such as permutation importance. Future work should investigate stronger xRFM hyperparameter tuning, more efficient AGOP approximation, improved categorical handling, and broader evaluation on sparse and multi-class tabular datasets.

\clearpage
